---
title: MACHINE LEARNING FOR SURVEY RESEARCH I
subtitle: Missing Data
author:
  - name: Marco R. Steenbergen
    orcid: 0000-0002-5709-5165
    email: steenbergen@ipz.uzh.ch
    affiliations: University of Zurich
date: last-modified
resource-path: ["..", "."]
format:
  clean-revealjs:
    revealjs-plugins:
      - tabset
    pdf-separate-fragments: true
    mermaid-format: svg
bibliography: /Users/prior/Resources/References/master.bib
self-contained: true
embed-resources: true
editor_options:
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

## Plan for the Day

| Element                | Content                                  |
|:-----------------------|:-----------------------------------------|
| **Part 1** (Morning)   | Missing data                             |
|                        | *Exercise 1 + Debrief*                   |
| **Part 2** (Afternoon) | Response styles                          |
|                        | *Exercise 2 + Debrief*                   |
| **Part 3**             | Multilevel ML for cross-national surveys |
|                        | *Exercise 3 + Debrief*                   |

## R Packages for this Lecture

Beyond packages you have installed already, please add

- `mice` for missing data imputation with chained equations
- `miceRanger` for implementing multiple imputation
- `missForest` for single RF-based imputations
- `tidymodels` for imputation inside a recipe
- `sbart` for sequential BART imputation --- theory only, the package is
  no longer maintained (see the caveat at the end)

# Basics of Missing Data {background-color="#40666e"}

## The Bane of the Social Scientist's Existence

- Missing data are a common occurrence in social science data analysis.
- They are common in survey research due to **non-response**.
- In archival work, they occur as well because reporting is not always
  complete.

## Missing Data Patterns {.smaller}

Two different questions, often conflated:

- **How much** is missing, and **where**? --- the plots on the next
  slide.
- **In what configuration** are the holes arranged? --- the *pattern*.

| Pattern | Shape | Consequence |
|:------------------|:---------------------|:------------------------------|
| **Monotone** | The variables can be ordered so that if $X_j$ is missing, everything after it is too --- classic panel attrition | Can be imputed **sequentially**, one variable at a time, in a single pass |
| **Non-monotone** | Holes scattered arbitrarily; no such ordering exists | We must **cycle** repeatedly through the variables until things settle |

::: callout-note
This is why MICE is called *chained* equations, and why it needs `maxit`
iterations at all. On a monotone pattern the chain would converge in one
sweep. Survey and archival data are almost always non-monotone.
:::

## Reasons for Missing Data

:::: panel-tabset
## Data

```{r}
#| echo: true
#| message: false
# We create a subset of the ESS Round 11 for Switzerland
# See the codebook (on GitHub) for explanations
library(haven)
library(tidyverse)
ess_raw <-  read_sav("../data/ESS11e04_2.sav", user_na = TRUE)
ess_ch <- ess_raw |>
  filter(cntry == "CH") |>
  select(trstprl, trstlgl, trstplc, trstplt, trstprt, trstep,
         trstun, hinctnta, lrscale, agea, gndr, eisced,
         domicil, hincfel, mnactic, rlgdgr, brncntr)
```

## Reasons I

```{r}
# We look at hinctnta, which has quite a few missing values
# 1-10 are valid; 77 = "Refusal", 88 = "Don't Know"
table(ess_ch$hinctnta)
```

## Reasons II

- There may be different reasons for the two missing responses.

- Refusal = the person does not want to reveal their income; missingness
  may be due to the value of the variable.

- DK = the person simply cannot remember their income; less likely to be
  due to the value of the income itself.

## Comment

::: callout-warning
## Beware

In a moment we will set all missing values to `NA`, losing all
distinctions between reasons why a response may be missing. Do this only
after you have gauged the various forms of missingness.
:::
::::

## Missing Value Patterns across Variables and Cases

::: panel-tabset
## Variablesgg_m

```{r}
#| echo: true
#| eval: false
library(naniar)
ess_ch <- ess_ch |>
  mutate(across(where(is.numeric), ~replace(.x, .x %in% c(66, 77, 88, 999, 999), NA)),
         brncntr = ifelse(brncntr==7, NA, brncntr),
         eisced = ifelse(eisced==0, NA, eisced)) 
gg_miss_var(ess_ch, show_pct = TRUE)
```

## Plot I

```{r}
#| message: false
library(naniar)
ess_ch <- ess_ch |>
  mutate(across(where(is.numeric), ~replace(.x, .x %in% c(66, 77, 88, 999), NA)),
         brncntr = ifelse(brncntr==7, NA, brncntr),
         eisced = ifelse(eisced==0, NA, eisced)) 
gg_miss_var(ess_ch, show_pct = TRUE)
```

## Cases

```{r}
#| echo: true
#| eval: false
gg_miss_case(ess_ch, order_cases = TRUE)
```

## Plot II

```{r}
gg_miss_case(ess_ch, order_cases = TRUE)
```

## Conclusions

- **By variable:** income (`hinctnta`) leads at 14.4%, then trust in the
  European Parliament (8.5%) and the UN (7.9%). Gender has none. Nothing
  here looks alarming.

- **By case:** only **976 of 1384** respondents (70.5%) are complete.
  The holes are spread across *different* people, so a per-variable rate
  of at most 14.4% compounds into a 29.5% loss for any analysis using
  all 17.
:::

## Missing Data Mechanisms

- Missing data mechanisms describe how the missing data were generated.

- Let $R$ indicate whether a value is observed, and split the data into
  what we see, $Y_{obs}$, and what we do not, $Y_{mis}$
  [@rubin1976Inference].

- Then:

  | Mechanism | Formally | In Words |
  |------------------------|------------------------|------------------------|
  | [**MCAR**]{.alert} | $P(R \mid Y_{obs}, Y_{mis}) = P(R)$ | Missingness is unrelated to anything, seen or unseen |
  | [**MAR**]{.alert} | $P(R \mid Y_{obs}, Y_{mis}) = P(R \mid Y_{obs})$ | Missingness depends **only on what we observe** |
  | [**NMAR**]{.alert} | $P(R)$ still depends on $Y_{mis}$ given $Y_{obs}$ | Missingness depends on the **unseen value itself** |

## Mechanisms and Reasons

- On income, "don't know" might be considered MAR.

- However, "refusal" might be more consistentr with MNAR.

## Why Mechanisms Matter

- MCAR does not introduce bias, but can introduce inefficiency. Testable
  but rare in social science research.

- MAR introduces both bias and inefficiency. However, can be treated
  with imputation. Not testable.

- MNAR also introduces bias and inefficiency. Here, imputation is
  ineffective and other solutions are required (e.g., the Heckman sample
  selection model).

# Traditional Approaches to Missing Data {background-color="#40666e"}

## Listwise Deletion

- A.k.a. complete case analysis.

- We drop all cases that have any missingness on the variables of
  interest.

- Unbiased under MCAR; generally biased for other mechanisms.

- Exception:

  - In a regression, it gives consistent estimates whenever missingness
    is conditionally independent of the *outcome* given the covariates.
  - That includes a class of MNAR mechanisms in which a covariate's
    missingness depends on its own value [@bartlett2014Improving;
    @xu2016Sequential].

## Hot-Deck Imputation

- Assuming MAR, we substitute the value of the nearest neighbor in terms
  of a set of covariates [@ford1983Overview].

- Akin to the k nearest neighbors (kNN) algorithm in ML.

## Parametric Approaches

- EM algorithm [@dempster1977Maximum].

- Bayesian imputation [@rubin1987Multiple].

- MICE = Multiple Imputation with Chained Equations
  [@vanbuuren2018Flexible].

  - Our focus because it is widely used.

# MICE {background-color="#40666e"}

## Mice Steps in the Imputation Stage

1.  We build a univariate imputation model for each incomplete variable,
    appropriate for its type.

2.  We cycle through the variables `maxit` times, each pass conditioning
    on the current imputed values of all the others.

3.  Each imputed value is drawn **with uncertainty**. Observed data are
    never overwritten.

4.  We repeat the process $M$ times.

    ::: notes
    \(1\) This is why MICE is called *chained* equations, and why it
    needs `maxit` iterations at all. On a monotone pattern the chain
    would converge in one sweep. Survey and archival data are almost
    always non-monotone. (2) Add noise to the prediction" describes
    `norm`, the Bayesian linear regression method. The **default for
    numeric variables is `pmm`**, predictive mean matching, which
    instead *borrows an observed value* from a donor whose predicted
    value is close. That is why `pmm` can never impute an impossible
    value --- no 4.7 on an integer 0--10 trust scale, no negative income
    decile.
    :::

## Mice Steps for Analysis

1.  We analyze each of the $m$ imputed data sets separately.

2.  We pool by Rubin's rules: average the estimates, and inflate the
    variance using both within- and between-data set variation:

    - Let $V_m$ denote the variance of an estimator using imputation
      data $m$.
    - The within-variance is the average:
      $\bar{V} = \frac{1}{M} \sum_m V_m$. This is the variance you would
      have with complete data.
    - The between-variance captures the variance between the estimators:
      $B = \frac{1}{M - 1} \sum_m (\hat{\theta}_m - \bar{\theta})^2$.
    - The total variance is
      $$T = \bar{U} + \left( 1 + \frac{1}{M} \right) \cdot B$$

::: notes
$1/M$ is a finite imputation correction that goes to 0 as
$M \rightarrow \infty$.
:::

## The Imputation Model

- The imputation model must be **at least as rich** as the analysis
  model: it has to contain the outcome, every predictor, and any
  interaction or nonlinearity you intend to estimate
  [@vanbuuren2018Flexible].

- Leaving the outcome out of the imputation model biases the very
  coefficients you are trying to estimate—toward zero.

- **Auxiliary variables** that predict either the missing values or the
  missingness itself belong in the imputation model even when they are
  absent from the analysis model. They make MAR more plausible.

## How Many Imputations

- @rubin1987Multiple proposed $2 \leq M \leq 10$ and that suffices for
  the relative efficiency of the point estimate.

- If the goal is to have a standard error that does not move much under
  re-imputation, then other rules apply.

- Let $\gamma$ be the **fraction of missing information**. Then

  - @white2011Multiple propose $M = 100 \gamma$.
  - @vonhippel2020How proposes
    $M = 1 + 0.5 \cdot \left( \frac{\gamma}{\text{CV}} \right)^2$, where
    CV = the coefficient of variation. Commonly, CV is set at 0.05.

## How Many Imputations Cont'd

- Define the **relative variance increase** as the ratio of the missing
  to the complete data part:

  | $$r = \frac{(1 + 1/M) \cdot B}{\bar{U}}$$

- The fraction of missing information is:

  $$\gamma = \frac{r + \frac{2}{\nu + 3}}{r + 1}$$

  where $\nu$ are the degrees of freedom.

## How Many Imputations Cont'd

| Variable   | Fraction Missing | $\bar{U}$ |      $B$ |      $T$ |   $r$ | $\gamma$ |
|------------|-----------------:|----------:|---------:|---------:|------:|---------:|
| Income     |             14.4 |   0005503 | 0.000138 | 0.000702 | 0.276 |    0.226 |
| Left-Right |              7.2 |  0.000641 | 0.000165 | 0.000822 | 0.283 |    0.230 |

Per @vonhippel2020How,
$M = 1 + 0.5 \cdot \left( \frac{0.230}{0.05} \right)^2 \approx 12$.

## MICE with ESS

::: panel-tabset
### Recoding

```{r}
#| echo: true
# An absolute mess because of our use of haven---it adds labels that other packages do not
# understand. Can be bypassed if you do not care about # reasons and use rio instead.
num_vars <- c("trstprl","trstlgl","trstplc","trstplt","trstprt","trstep",
              "trstun","hinctnta","lrscale","agea","rlgdgr")
to_num <- function(x) replace(zap_labels(x), is.na(x), NA_real_)
work <- ess_ch |>
  mutate(
    across(all_of(num_vars), to_num),
    
    eisced = case_match(to_num(eisced),
                        1 ~ "I", 2 ~ "II", 3 ~ "IIIb", 4 ~ "IIIa",
                        5 ~ "IV", 6 ~ "V1", 7 ~ "V2", 55 ~ "Other",
                        .ptype = factor(levels = c("I","II","IIIb","IIIa","IV","V1","V2","Other"))),
    
    hincfel = case_match(to_num(hincfel),
                         1 ~ "Comfortable", 2 ~ "Cope", 3 ~ "Difficult", 4 ~ "Very Difficult",
                         .ptype = factor(levels = c("Comfortable","Cope","Difficult","Very Difficult"),
                                         ordered = TRUE)),
    
    domicil = case_match(to_num(domicil),
                         1 ~ "Big City", 2 ~ "Suburb", 3 ~ "Town", 4 ~ "Country Village", 5 ~ "Farm",
                         .ptype = factor(levels = c("Big City","Suburb","Town",
                                                    "Country Village","Farm"))),
    
    mnactic = case_match(to_num(mnactic),
                         1 ~ "Paid Work", 2 ~ "Education", 3 ~ "Unemployed and Looking",
                         4 ~ "Unemployed Not Looking", 5 ~ "Disabled", 6 ~ "Retired",
                         7 ~ "Military Service", 8 ~ "Housework", 9 ~ "Other",
                         .ptype = factor(levels = c("Paid Work","Education","Unemployed and Looking",
                                                    "Unemployed Not Looking","Disabled","Retired",
                                                    "Military Service","Housework","Other"))),
    
    brncntr = case_match(to_num(brncntr),
                         1 ~ "Yes", 2 ~ "No", .ptype = factor(levels = c("Yes","No"))),
    
    gndr = case_match(to_num(gndr),
                      1 ~ "Male", 2 ~ "Female", .ptype = factor(levels = c("Male","Female")))
  )
```

### Imputation

```{r}
#| echo: true
library(mice)
set.seed(2026)
imp <- mice(work,
            m       = 10,    
            maxit   = 10,
            printFlag = FALSE)
```

### Diagnostics

```{r}
#| echo: true
#| eval: false
# imputed (red) and observed (blue) distributions should look plausible.
densityplot(imp, ~ trstprl + hinctnta + lrscale + hincfel )
```

### Plot

```{r}
densityplot(imp, ~ trstprl + hinctnta + lrscale + hincfel)
```

### Analysis

```{r}
#| echo: true
fit <- with(imp, lm(trstprl ~ hinctnta + lrscale + hincfel + agea + 
                      brncntr + domicil))
pooled <- pool(fit)
summary(pooled, conf.int = TRUE)
```
:::

# Imputation Using Machine Learning {background-color="#40666e"}

## Why Machine Learning for Missing Data?

- Non-parametric models.

- Capable of learning complex interplay of covariates, as well as
  nonlinear patterns.

- Tend to work well even when the amount of missing data is large.

## Theory of Missing Forests

- @stekhoven2012MissForest proposed imputation using random forests.

- Training proceeds as follows:

  1.  Initialize by making simple baseline guesses (mean or mode) and by
      sorting on the amount of missingness $\Rightarrow$ start with
      least missing data.

  2.  \(a\) Divide the data in to observed and missing entries on a
      variable. (b) divide all other variables into predictor rows
      corresponding to the observed and missing parts of the target. (c)
      Train the forests on the observed data. (d) Predsict and
      overwrite.

  3.  Stop when matrix differences stabilize over iterations.

## Random Forest Imputation

:::: panel-tabset
## Setup

```{r}
#| echo: true
#| message: false
library(missForest)
# Tree-based imputers handle ordered factors awkwardly, so we drop the
# ordering. missForest also wants a plain data frame, not a tibble.
work_ml <- work |>
  mutate(hincfel = factor(hincfel, ordered = FALSE)) |>
  as.data.frame()
```

## Syntax

```{r}
#| echo: true
#| eval: false
set.seed(2026)
ess_rf_imp <- missForest(work_ml)
# NRMSE for the numeric block, PFC (proportion falsely classified) for the factors
ess_rf_imp$OOBerror
```

## Results

```{r}
#| message: false
set.seed(2026)
ess_rf_imp <- missForest(work_ml)
ess_rf_imp$OOBerror
```

## Completed Data

```{r}
#| echo: true
ess_rf_complete <- ess_rf_imp$ximp
sum(is.na(ess_rf_complete))          # 0 --- every hole is filled
summary(ess_rf_complete$hinctnta)    # note: no longer integer-valued
```

::: callout-note
The forest predicts a *conditional mean*, so imputed income deciles come
out as 6.38, not 6. Contrast `pmm` in `mice`, which borrows an observed
value from a donor and therefore cannot produce an impossible one.
:::
::::

## The Drawback

- Only one imputed data set is returned: no capture of uncertainty.

- @doove2014Recursive and @shah2014Comparison proposed multiple
  imputation with RFs.

- Currently, the fastest implementation is `miceRanger`.

## Using `miceRanger`

:::: panel-tabset
## Impute

```{r}
#| echo: true
#| message: false
library(miceRanger)
set.seed(2026)
mr <- miceRanger(work_ml,
                 m       = 12,       # per the von Hippel calculation earlier
                 maxiter = 5,
                 valueSelector       = "meanMatch",  # predictive mean matching on RF predictions
                 meanMatchCandidates = 15,
                 verbose = FALSE)
```

::: notes
This is the slowest chunk in the deck: 12 x 5 x (number of incomplete
variables) forests. Raise `maxiter` to 10 to match the `mice` call if
you have the time.
:::

## Checks

```{r}
#| echo: true
#| eval: false
# Ideally, you should run these checks to assess plausibility and
# convergence. For the latter, jaggedness is OK--trending is not.
plotDistributions(mr, vars = c("hinctnta", "lrscale", "trstep", "trstun"))
plotVarConvergence(mr)   # means/SDs of imputed values across iterations
plotModelError(mr)       # OOB R^2 per variable --- how well the forests predict
plotVarImportance(mr)    # which variables drive each imputation model
```

## Model

```{r}
#| echo: true
dat_list <- completeData(mr)
fits <- lapply(dat_list, function(d)
  lm(trstprl ~ hinctnta + lrscale + hincfel + agea + brncntr + domicil,
     data = d))
pooled_rf <- mice::pool(fits)
summary(pooled_rf, conf.int = TRUE)
```
::::

## Comparing the Three Routes {.smaller}

|   | `mice` (pmm) | `missForest` | `miceRanger` |
|:-----------------|:-----------------|:-----------------|:-----------------|
| Data sets returned | $M$ | 1 | $M$ |
| Uncertainty | Rubin's rules | none | Rubin's rules |
| Nonlinearity, interactions | only if you specify them | learned | learned |
| Imputed values | always plausible (donor) | conditional means | plausible (donor) |
| Cost | low | low | high |

::: callout-note
Because `hincfel` is now an unordered factor, its contrasts differ from
the `mice` run above. Compare the numeric coefficients, not the
`hincfel` block, when you look at the two pooled tables side by side.
:::

# Imputation Inside a Recipe {background-color="#40666e"}

## Why Bother?

- Everything so far imputes the **whole** data set once and then models
  it. In a predictive workflow that is a leak: the imputation model has
  already seen the rows you are about to hold out.

- A `recipes` step is estimated on the *analysis* part of each resample
  and merely *applied* to the *assessment* part. The imputation is
  re-learned for every fold, exactly like any other preprocessing.

- It also travels with the fitted model: calling `predict()` on
  genuinely new data that has holes in it just works.

## What `recipes` Gives You {.smaller}

| Step | Learner | Comment |
|:-------------------|:-------------------|:-------------------------------|
| `step_impute_bag()` | bagged CART (`ipred`) | The RF-like option: an ensemble of trees, mixed types, interactions for free |
| `step_impute_knn()` | Gower-distance kNN | Hot-deck imputation, essentially |
| `step_impute_linear()` | `lm` | Numeric predictors only |
| `step_impute_median()`, `step_impute_mode()` | marginal | Fast baseline; ignores everything else |

::: callout-note
There is no `step_impute_rf()`. Bagged trees are missForest minus the
random subsetting of predictors at each split --- the same idea, a
little less decorrelated.
:::

## A Recipe on the ESS

:::: panel-tabset
## Split

```{r}
#| echo: true
#| message: false
library(tidymodels)
tidymodels_prefer()

# recipes imputes predictors, never the outcome: drop those rows first
ess_ml <- work_ml |> filter(!is.na(trstprl))

set.seed(2026)
ess_split <- initial_split(ess_ml, prop = 0.8)
ess_train <- training(ess_split)
ess_test  <- testing(ess_split)
```

## Recipe

```{r}
#| echo: true
imp_rec <- recipe(trstprl ~ ., data = ess_train) |>
  step_impute_bag(all_predictors(),
                  impute_with = imp_vars(all_predictors()),
                  trees       = 25,
                  seed_val    = 2026) |>
  step_dummy(all_nominal_predictors()) |>
  step_normalize(all_numeric_predictors())
```

::: callout-warning
Order matters. Impute **before** `step_dummy()` and `step_normalize()`.
Imputing after dummying means imputing 0/1 indicator columns, which is
nonsense.
:::

## Check

```{r}
#| echo: true
imp_prep <- prep(imp_rec)
baked <- bake(imp_prep, new_data = NULL)
sum(is.na(baked))     # 0
```

## Workflow

```{r}
#| echo: true
#| message: false
rf_spec <- rand_forest(trees = 500) |>
  set_engine("ranger") |>
  set_mode("regression")

imp_flow <- workflow() |>
  add_recipe(imp_rec) |>
  add_model(rf_spec)

set.seed(2026)
imp_fit <- fit(imp_flow, data = ess_train)

imp_fit |>
  predict(ess_test) |>
  bind_cols(ess_test) |>
  metrics(truth = trstprl, estimate = .pred)
```
::::

## The Catch {.smaller}

::: callout-important
## A recipe does single imputation

`step_impute_bag()` returns **one** completed data set per fold. There
is no between-imputation variance, so there is nothing to pool by
Rubin's rules and the standard errors on any coefficient you read off
are too small.
:::

- That is perfectly fine when the target is **predictive performance**:
  RMSE on the assessment set already accounts for imputation error,
  because the imputation was fit without seeing those rows.

- It is *not* fine when the target is **inference** on a coefficient.
  For that, impute outside the workflow with `mice` or `miceRanger`, run
  the workflow $M$ times, and pool by hand.

- Cost: `step_impute_bag()` is the most expensive step in `recipes`.
  With `v = 10` folds you fit 25 trees x (incomplete variables) x 10.

- Alternative: skip imputation entirely and use a learner that eats
  missing values --- `rpart` surrogate splits, or XGBoost and BART,
  which treat missingness as a value in its own right when choosing a
  split.

# Sequential BART {background-color="#40666e"}

## The Problem It Solves

- MICE has no joint distribution behind it. The univariate conditionals
  you specify need not be **compatible** --- there may exist no joint
  distribution having all of them as conditionals [@xu2016Sequential].

- A second issue is **uncongeniality**: the imputation model and the
  analysis model can imply different things about the data, and Rubin's
  rules are not guaranteed to behave.

- The parametric fixes (sequential regression on a monotone
  factorization) restore compatibility but throw away flexibility: they
  cannot capture nonlinearity or interactions.

## The Sequential Factorization

- @xu2016Sequential factor the joint distribution of the incomplete
  covariates **exactly**:

  $$p(X_1, X_2, \ldots, X_p) = p(X_1) \prod_{j = 2}^{p} p(X_j \mid X_1, \ldots, X_{j-1})$$

- Each univariate conditional on the right is then modeled with BART
  [@chipman2010Bart] --- the sum-of-trees machine from this morning.

- Because the factorization is an identity, the conditionals **are**
  compatible by construction. There is a real joint model, unlike in
  MICE.

- Flexibility is retained: each conditional is nonparametric, so
  nonlinearity and interactions are learned rather than declared.

## The Machinery

1.  Missing entries are treated as **parameters** and handled by **data
    augmentation**: at each MCMC sweep, draw the missing values from
    their full conditional given the current trees.

2.  Update the BART ensemble for each conditional given the now-complete
    data.

3.  Because everything sits inside one MCMC run, all conditionals are
    sampled **simultaneously** rather than cycled through as in MICE.
    There is no `maxit` and no separate convergence-of-the-chain
    diagnostic per variable.

4.  The $M$ imputations are simply $M$ thinned draws from the posterior.
    Rubin's rules then apply as usual.

::: notes
This is the substantive difference from MICE. MICE alternates between
conditionals that may not cohere; sequential BART draws from a single
posterior in which the conditionals cohere by construction.
:::

## Does It Help?

- In @xu2016Sequential's simulations with nonlinear covariate
  relationships, sequential BART had bias near zero for every covariate,
  while MICE, MICE-CART, and sequential parametric imputation were
  biased on at least some.

- Confidence interval coverage was close to nominal for sequential BART
  and poor for the competitors, and the intervals were also the
  **narrowest** --- better coverage at lower cost.

- It handles MAR whether or not missingness depends on the response.

- It was designed for the many-covariates, moderate-$n$ setting of
  electronic health records --- which is more or less the setting of a
  cross-national survey.

## Caveats

- The ordering of the sequential factorization is arbitrary and the
  answer is not perfectly invariant to it.

- Categorical and binary covariates need a probit / latent-variable
  layer; continuous covariates are the natural case.

- Cost. One BART ensemble per incomplete variable per sweep, over
  thousands of sweeps. Orders of magnitude slower than `miceRanger`.

## A Warning About the Package

::: callout-warning
## `sbart` is no longer maintained

The reference implementation (`seqBART()` in the `sbart` package) has
been archived and no longer builds against current R. Treat this section
as **theory you should know**, not as a tool for Exercise 1.
:::

If you want the spirit of it in maintained code today:

- `bartMachine(use_missing_data = TRUE)` --- does not impute at all, but
  lets missingness itself serve as a splitting criterion, so the holes
  are carried straight into the model.

- `mice(method = "cart")` or `miceRanger` --- keeps the
  chained-equations machinery, buys back the tree flexibility, gives up
  the compatibility guarantee.

## References
